Complex and informative dependencies propagate throughout ecology and can be revealed using copulas

### 2019-05-13 Paper.Rmd
still editing

---
title: "Complex, informative, and previously ignored dependencies propogate throughout ecology and can be revealed using copulas"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

@@

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of density functions of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for OH), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
density function of a normal copula (methods for OH) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band (diagonal lines
show two bands) can be computed 
for any band (methods for OH) to describe how dependence
varies from one part of the two distributions to another.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in finance, neuroscience, bioinformatics and other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure only.
Mathematically, if $(X,Y)$ is a continuous random
variable and the cumulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in $[0,1] \times [0,1]$, and has 
uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}C-D are superimposed on those panels. 
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A principal benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
characteristics (Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for OH).

The goal of this paper is to illustrate the potential for 
broader application of copulas in ecology. 
Our results provide initial support for the following 
overall hypothesis: environmental, ecological and evolutionary processes 
generate and transmit copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding and for 
applications across ecology; analyzing copula structure 
provides more detailed tests of ecological theories and more revealing 
desciptions of ecological datasets. This is a very broad hypothesis, and 
even the substantial empirical and theoretical
evidence we assembled here has only begun to test it. So we propose the hypothesis
principally for further evaluation by the research community.
The overall hypothesis is examined by assessing several more specific 
hypotheses, described below and outlined in Fig. 
\ref{MasterFigure}, all of which bear upon the overall hypothesis.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. CaH="causal hypothesis" (a hypothesis about 
causes of copula structure), CoH="consequential hypothesis" (a hypothesis about 
consequences of copula structure) and OH="observational hypothesis" (a hypothesis about 
copula structure of data).\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portion of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques. We therefore frame "causal hypothesis" H1: 
copulas of dependencies between environmental and 
ecological variables will sensitively reveal 
nonlinear, asymmetric and threshold effects for population, community, 
and ecosystem functioning variables. Copulas between 
environmental and ecological variables
will commonly show tail dependence. This is labeled CaH1 in Fig. \ref{MasterFigure};
we use the term "causal hypothesis" because it is about a prospective cause of 
copula structure.

A second reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
X and Y are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If X is plotted on the horizontal axis
and Y on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed; 
whereas when X is rare, Y
may be abundant, or may also be relatively rare due to 
limiting factors other than X. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Y
is dominant. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
information can help support or refute alternative causal hypotheses. We
frame causal hypothesis H2 (CaH2 on Fig. \ref{MasterFigure}): copulas of abundances
through time or space of two competing or otherwise interacting species will
reflect the asymmetries in the interaction.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of the spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017]. We therefore frame 
causal hypothesis 3 (CaH3 on Fig. \ref{MasterFigure}): Moran effects
transmit copula structure in spatial dependence from environmental to 
ecological variables, and hence copula structure and tail dependence will be 
common features of spatial synchrony in ecological variables. We will later
frame CaH4, which concerns character evolution.

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We therefore frame "consequential hypothesis" H1 (CoH1 on Fig. \ref{MasterFigure}):
details, revealed by copulas, of the 
dependence between local quantities can influence the spatial 
mean or sum, and this may influence higher organizational levels in 
ecology and human concerns. We use the term "consequential hypothesis"
because it is about a prospective consequence of copula structure.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have much more benign effects. We will
later frame CoH2, about extinction risk, and CoH3, about Taylor's law.

Our causal and consequential hypotheses imply, or rely on, an "observational
hypothesis," OH, which we here frame: datasets in ecology will commonly have copula structure
distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure). They will commonly show tail dependence 
distinct from that of a normal distribution. In particular they will
show asymmetric tail dependence. We use the term "observational hypothesis"
because it pertains to what copula structures will be observed in data.

Because OH bears upon all the causal
and consequential hypotheses, we test OH first, in sections 2-4 of the paper.
We analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
Sections 5 and 6 test the causal hypotheses, and sections 7 and 8 test the
consequential hypotheses. Section 9 is the Discussion.
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, these concepts have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula desription of dependence will be substantial.

@@

# Background, concepts and methods for OH: Detecting copula structure in data \label{Methods}

To make results interpretable
to unfamiliar readers, we first introduce the copula families we used, and 
measures of tail dependence. 
Conceptually, a bivariate copula is a probability density function (pdf)
on the unit square with uniform maginals. 
The copula is actually defined as the corresponding 
cdf rather than as the pdf,
but both functions contain essentially the same information. 
The normal copula family (henceforth abreviated N) is a one-parameter
family of copulas, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag3}A-B for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than a normal
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}G-H) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}G) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}H)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a variety of 
positive lower-tail dependence values 
(Clayton, survival-Gumbel, survival-Joe, survival-BB6, henceforth
abbreviated C, SG, SJ, SBB6) and upper-tail dependence values
(survival-Clayton, Gumbel, Joe, BB6, henceforth abbreviated 
SC, G, J, BB6). We also considered families that could exhibit positive upper- or lower-tail
dependence, or both, depending on parameters (BB1, survival-BB1, BB7, survival-BB7, henceforth 
abbreviated BB1, SBB1, BB7, SBB7). 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence. We also used two families that were asymmetric in their non-tail parts, 
but that had zero tail dependence (BB8, survival-BB8, abbreviated BB8 and SBB8), and
a symmetric copula called 
the Frank copula (abbreviated F; Fig. \ref{fig_cop_pedag3}C-D). 
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R language. Visualizations of the copulas we used are in Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
See @nelsen2006_copula, @joe2014_dependence and the `VineCopula` documentation
for more details.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig3.pdf}
\caption{Log-transformed pdfs (A, C) and samples from (B, D) example copulas.
Shown are the normal (N) copula with parameter $0.8$ and the Frank (F) copula with 
parameter $10$. See also Fig. \ref{fig_cop_pedag1}E-H for the Clayton (C) and  
survival-Clayton (SC) copulas with parameter 5. Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}
have visualizations of all the copulas we used.\label{fig_cop_pedag3}}
\end{center}
\end{figure}

We addressed OH principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using the `BiCopEst` function in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The method performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of OH was answered for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part of OH was answered by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). The third part of OH
was answered by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we carried out the bivariate analysis
for all possible pairwise combinations of locations, and compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of pairs showing stronger lower-tail dependence out of 
the total number of non-independent location pairs considered.

>>>>

---
title: "Complex and informative dependencies propagate throughout ecology and can be revealed using copulas"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

Complex, informative, and previously ignored dependencies propagate throughout ecology and can be revealed using copulas

The complexity of ecological relationships is revealed by copulas

The true complexity of ecological relationships is revealed by copulas

The true complexity of ecological relationships is only revealed by copulas

@@

Copulas reveal complex and informative dependence
structures between variables across ecological disciplines

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of density functions of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for the observational hypothesis), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
density function of a normal copula (methods for the observational hypothesis) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band (diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band) can be computed 
for any band (methods for the observational hypothesis) to describe how dependence
varies from one part of the two distributions to another.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in finance, neuroscience, bioinformatics and other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure only.
Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}C-D are superimposed on those panels. 
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A principal benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
characteristics (Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for the observational hypothesis).

The goal of this paper is to illustrate the potential for 
broader application of copulas in ecology. 
Our results provide initial support for the following 
overall hypothesis: environmental, ecological and evolutionary processes 
generate and transmit copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. This is a broad hypothesis, and 
even the substantial empirical and theoretical
evidence we present here has only begun to test it. So we hope our work stimulates further
evaluation of the hypothesis by the research community.
Our overall hypothesis is examined by assessing several more specific 
hypotheses, described below and outlined in Fig. 
\ref{MasterFigure}, all of which bear upon the overall hypothesis.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. "Causal hypotheses" (i.e., hypotheses about 
causes of copula structure) are labeled with letters A-D. "Consequential 
hypothesis" (i.e., hypothesis about 
consequences of copula structure) are labeled with letters X-Z. The "observational 
hypothesis" (i.e., our hypothesis about 
copula structure of data) is labeled with O.\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques. We therefore frame "causal hypothesis" A 
(labeled in Fig. \ref{MasterFigure}): 
copulas of dependencies between environmental and 
ecological variables will sensitively reveal 
nonlinear, asymmetric and threshold effects for population, community, 
and ecosystem functioning variables; copulas between 
environmental and ecological variables
will commonly show tail dependence; such tail dependence will lead to 
tail dependence across space in ecological variables.

A second reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of the spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017]. We therefore frame 
causal hypothesis B (again labeled in Fig. \ref{MasterFigure}): Moran effects
transmit copula structure in spatial dependence from environmental to 
ecological variables, and hence copula structure and tail dependence will be 
common features of spatial synchrony in ecological variables.

A third reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Sp2
is dominant. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
information can help support or refute alternative causal hypotheses. We
frame causal hypothesis C (Fig. \ref{MasterFigure}): copulas of abundances
through time or space of two competing or otherwise interacting species will
reflect any asymmetries in the interaction. We will later
frame causal hypothesis D, which concerns character evolution.

Another reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We therefore frame "consequential hypothesis" X (labeled on Fig. \ref{MasterFigure}):
details, revealed by copulas, of the 
dependence between local quantities can influence the spatial 
mean or sum, and this may influence higher organizational levels in 
ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects. We will
later frame consequential hypotheses Y, about extinction risk, and Z, about Taylor's law.

Our causal and consequential hypotheses imply, or rely on, an "observational
hypothesis" (O in Fig. \ref{MasteFigure}), which we here frame: datasets in 
ecology will commonly have copula structure
distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure); they will commonly show tail dependence 
distinct from that of a normal distribution; in particular they will
show asymmetric tail dependence.

Because the observational hypothesis bears upon all the causal
and consequential hypotheses, we test it first, in sections 2-4 of the paper.
We analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
Sections 5 and 6 test the causal hypotheses, and sections 7 and 8 test the
consequential hypotheses. Section 9 is the Discussion.
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, these concepts have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula desription of dependence will be substantial.

@@

# Background, concepts and methods for testing the observational hypothesis \label{Methods}

We introduce the copula families we used, and 
measures of tail dependence. 
Conceptually, a bivariate copula is a pdf
on the unit square with uniform maginals. 
The copula is actually defined instead as the corresponding 
cummulative dstribution function (cdf), but both functions contain the same information.
If $(X,Y)$ is a continuous random
variable and the cdfs of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in $[0,1] \times [0,1]$, and has 
uniform marginal distributions. 
The normal copula family (henceforth abreviated N) is a one-parameter
family of copulas, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several one- and two-parameter families of copulas exhibiting a variety of 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the `copula` and `VineCopula` packages for the 
R language. Visualizations of the copulas we used are in 
Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence and the `VineCopula` documentation
for further background.

We tested the observational hypothesis (O in Fig. \ref{MasterFigure}) 
principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using the `BiCopEst` function in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The method performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of the observational hypothesis was tested for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part was tested by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part of the observational hypothesis
was tested by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we carried out the bivariate analysis
for all possible pairwise combinations of locations, and compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of pairs showing stronger lower-tail dependence out of 
the total number of non-independent location pairs considered.

In addition to our model selection approach, we also used nonparametric approaches to 
provide greater 
confidence in our tests of the observational hypothesis. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All the variables we study were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association
between the bounds.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
for points between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.
